### 2020-06-10 chapters/sota.tex
Outlier detection method taxonomy

%Probably say we analyze outlier methods and why aggregations? relate to the approaches thought of to solve the problem and the sections

@@

The \textit{input data} axis refers to the type of time-series the method uses as input. According to this criterion, an outlier detection method is said to work with a univariate time-series or a multivariate one. A univariate time-series refers to a time-series consisting of a list of values generated by a single time-dependent variable throughout time. To quote the authors, a univariate time-series is \textit{"an ordered set of real-valued observations, where each observation is recorded at a specific time"}. On the other hand, a multivariate time-series has more than one time-dependent variable. Most importantly, each variable is not only time-dependant but may also have dependencies on other variables. With that in mind, the authors define a multivariate time-series as \textit{"an ordered set of k-dimensional vectors, each of which is recorded at a specific time"}. Each \textit{k-dimensional} vector consists of \textit{k}  time-dependent variables, possibly correlated.

@@

The univariate point outlier detection methods based on models are the most common. For each timestamp of the time-series, the models compute the expected value for the time-dependent variable and compare it with the observed value. If the difference between the expected value and the observed one is above a certain threshold the data point at the current timestamp is considered an outlier. Mathematically, given the expected value $e\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold of $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if $|$x\textsubscript{t}$-$e\textsubscript{t}$|>\alpha$. Additionally, the authors split the model-based category into estimation and prediction models. The main difference is that estimation model-based methods use past and future data while prediction model-based methods use only past data and infer current and future values. Because of this, prediction model-based outlier detection methods can be used in real-time to process streaming time-series, while estimation based methods can't.

Density-based univariate point outlier detection methods rely on clustering of data points and neighbor counting. These methods consider a datum an outlier if it has less than $\alpha$ neighbors. Distance between two points is usually measured using Euclidean distance. Two points are said to be neighbors if they are close to each other --- \textit{i.e.} the distance between them is less than or equal to a given threshold $\alpha$.

@@

HTM\\
point outlier, model based, prediction\\
black box ML algorithm\\
explain properties\\
use what authors say in their review\\

DeepAnT\\
point outlier, model based, prediction\\
black box DL algorithm\\
explain properties\\
use what authors say in their review\\
\fi

\subsection*{Univariate subsequence outlier detection}
Similarly to point outlier detection methods, the authors split subsequence outlier detection methods into uni and multivariate ones \ref{fig:outlier-taxonomy}. Furthermore, the authors split univariate subsequence outlier detection methods into five different axes as seen in Figure \ref{fig:subsequence-outlier-taxonomy}.

@@

Methods in the discord detection category compare each subsequence in the time-series with the other subsequences within the time-series in order to find out the most unusual ones. Each subsequence is compared to the others using a distance metric, for instances the Euclidean distance. The brute-force way of accomplishing this requires a pairwise comparison of all the subsequences, effectively resulting in a quadratic time complexity. These methods find the most unusual subsequence in a time-series. However, as the authors put it, these type of methods \textit{"do not have a reference of normality or a threshold"} and hence they do not accurately report outliers.

Univariate subsequence outlier detection methods based on dissimilarity are identical to the previous discord detection based methods but use a reference of normality. The representations used for the subsequences and the reference of normality vary among methods in this class of outlier detection methods but they are based on comparison of the reference or normality subsequence with the ones being analyzed for outlier detection. Once again, the metric used to measure the distance between the reference period subsequence and the other ones depends entirely on the method and subsequence aggregation or representation. Finally, when the measured distance is above a certain threshold, these methods classify the subsequence as an outlier or an anomaly.

@@

>>>>

%Probably say we analyze outlier methods and why aggregations? relate to the approaches thought to solve the problem and the sections

@@

The \textit{input data} axis refers to the type of time-series the method uses as input. According to this criterion, an outlier detection method is said to work with a univariate time-series or a multivariate one. A univariate time-series refers to a time-series consisting of a list of values generated by a single time-dependent variable throughout time. To quote the authors, a univariate time-series is \textit{"an ordered set of real-valued observations, where each observation is recorded at a specific time"}. On the other hand, a multivariate time-series has more than one time-dependent variable. Most importantly, each variable is not only time-dependent but may also have dependencies on other variables. With that in mind, the authors define a multivariate time-series as \textit{"an ordered set of k-dimensional vectors, each of which is recorded at a specific time"}. Each \textit{k-dimensional} vector consists of \textit{k}  time-dependent variables, possibly correlated.

@@

The univariate point outlier detection methods based on models are the most common. For each timestamp of the time-series, the models compute the expected value for the time-dependent variable and compare it with the observed value. If the difference between the expected value and the observed one is above a certain threshold the data point at the current timestamp is considered an outlier. Mathematically, given the expected value $e\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if
\begin{equation*}
    |x\textsubscript{t}-e\textsubscript{t}|>\alpha
\end{equation*}
Additionally, the authors split the model-based category into estimation and prediction models. The main difference is that estimation model-based methods use past and future data while prediction model-based methods use only past data and infer current and future values. Because of this, prediction model-based outlier detection methods can be used in real-time to process streaming time-series, while estimation based methods can't.

Density-based univariate point outlier detection methods rely on the clustering of data points and neighbor counting. These methods consider a datum an outlier if it has less than $\alpha$ neighbors. Distance between two points is usually measured using Euclidean distance. Two points are said to be neighbors if they are close to each other --- \textit{i.e.} the distance between them is less than or equal to a given threshold $\alpha$.

@@

HTM\\
point outlier, model-based, prediction\\
black-box ML algorithm\\
explain properties\\
use what authors say in their review\\

DeepAnT\\
point outlier, model-based, prediction\\
black box DL algorithm\\
explain properties\\
use what authors say in their review\\
\fi

\subsection*{Univariate subsequence outlier detection}
Similarly to point outlier detection methods, the authors split subsequence outlier detection methods into uni and multivariate ones (Figure \ref{fig:outlier-taxonomy}). Furthermore, the authors split univariate subsequence outlier detection methods into five different axes as seen in Figure \ref{fig:subsequence-outlier-taxonomy}.

@@

Methods in the discord detection category compare each subsequence in the time-series with the other subsequences within the time-series to find out the most unusual ones. Each subsequence is compared to the others using a distance metric, for instance, the Euclidean distance. The brute-force way of accomplishing this requires a pairwise comparison of all the subsequences, effectively resulting in quadratic time complexity. These methods find the most unusual subsequence in a time-series. However, as the authors put it, these types of methods \textit{"do not have a reference of normality or a threshold"} and hence they do not accurately report outliers.

On the other hand, univariate subsequence outlier detection methods based on dissimilarity are identical to the previous discord detection based methods but use a reference of normality. The representations used for the subsequences and the reference of normality vary among methods in this class of outlier detection methods but they are based on comparison of the reference or normality subsequence with the ones being analyzed for outlier detection. Once again, the metric used to measure the distance between the reference period subsequence and the other ones depends entirely on the method and subsequence aggregation or representation. Finally, when the measured distance is above a certain threshold, these methods classify the subsequence as an outlier or an anomaly.

@@

Frequency-based univariate subsequence outlier detection methods also use a reference period and compute the frequency of each subsequence in the reference or normality period. Hence, a subsequence will be an outlier if it does not appear as frequently as expected in subsequent time periods.

Lastly, we have information theory based methods for univariate subsequence outlier detection, which according to the authors \textit{"are closely related to the frequency-based methods"}. This type of method not only measures the frequency of subsequences but relates it with the information carried by the subsequence. Mathematically, a subsequence \textit{S} will be an outlier if

\begin{equation*}
    I(S) \times F(S) > \alpha
\end{equation*}

with \textit{F(S)} as the frequency of subsequence S, \textit{I(S)} as the information carried by S and $\alpha$ as a given threshold. According to the authors, \textit{"I(S) is computed taking into account the number of times the symbols within S are repeated through the series"}. The authors explain that if \textit{F(S)} is large --- \textit{i.e.} S occurs frequently --- then \textit{I(S)} will be closer to 0.
